Fail Faster

### 2025-03-17 paper/paper.tex
wibbleg

In Muli-stage programming (also known more simply as staging), programs execute over the course of multiple stages,
with each stage producing the code to be run in the next.
This is an old idea, with roots going back to quasiquotation in LISP \cn{}, and picking up steam again in the 1990s with MetaML \cn{}.
For the purposes of this paper, we will primarily consider two stages: compile time and run time. Staged programs thus execute twice: once at compile time,
which produces more code, which is then compiled and run at run time.

@@

One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
While embedded DSLs are great \jwc{words}, they have well-known drawback: 
the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
In the literature, this effect is known as ``abstraction overhead'': the layers of abstraction that 
make eDSLs so usable are precisely what prevents them from being fast.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
\jwc{This previous paragraph is peak ``shitty first draft''.}
This recipe has been used to great effect across domains to stage eDSLs for stream processing \cn, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

Many languages have some degree of staging functionality, though support varies
from full native support (Scala 3 \cn, Haskell \cn) to experimental compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}, Kotlin \cn).
Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
See Section~\ref{subsection:other-langs} for a full discussion of just how portable these ideas are. \jwc{this sentence is a placeholder lol}

@@

In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
must allocate lists in the hot path of the generator. Even though these lists are often small ---
usually at most a few dozen elements in practice --- each allocation takes us closer to the next garbage collection.

This is an ideal opportunity to exercise another use of staging: compile-time specialization.
Since we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
can generate \emph{different code} depending on the number of generators in the union.
If we use weighted union on a compile-time list of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
we can emit code that picks between the generators without realizing the list at
run time.

>>>>

In Muli-stage programming (also known more simply as staging), programs execute
over the course of multiple stages, with each stage producing the code to be run
in the next.  This is an old idea in programming languages, with roots going
back to quasiquotation in LISP \cn{}.  For the purposes of this paper, we will
primarily consider two stages: compile time and run time. Staged programs thus
execute twice: once at compile time, which produces more code, which is then
compiled and run at run time.

@@

One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
While embedded DSLs are great \jwc{words}, they have well-known drawback: 
the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
In the literature, this effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
make eDSLs so usable are precisely what prevents them from being fast.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
\jwc{This previous paragraph is peak ``shitty first draft''.}
This recipe has been used to great effect across domains to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

Many languages have some degree of staging functionality, though support varies
from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
See Section~\ref{subsection:other-langs} for a full discussion of just how portable these ideas are. \jwc{this sentence is a placeholder lol}

@@

In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
often allocate lists in the hot path of the generator. Even though these lists are usually small ---
at most a few dozen elements in practice --- each allocation takes us closer to the next garbage collection.

This is an ideal opportunity to exercise another feature of staging: compile-time specialization.
Since we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
can generate \emph{different code} depending on the number of generators in the union.
If we use weighted union on a compile-time list of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
we can emit code that picks between the generators without realizing the list at
run time.
